Modelling the City of Melbourne Pedestrian Data

### 2017-09-30 presentation/final.Rmd
update - fixed slide titles

## Format of the data
#### Variables available:
- Time and Date
- Sensor ID/Location
- Hourly Pedestrian Count

```{r, echo=FALSE}
ped_data %>% as.tibble
```

---
```{r, include = FALSE, echo = FALSE, warning=FALSE, message=FALSE, cache = TRUE}
ped_data$Sensor_ID <- ped_data$Sensor
ped_data$Hourly_Counts <- ped_data$Count
ped_data$Date_Time <- ymd_h(paste(ped_data$Date, ped_data$Time, sep = " "))
ped_data$Day <- lubridate::wday(ped_data$Date_Time, label = TRUE, abbr = F)
ped_data$Month <- lubridate::month(ped_data$Date_Time, label = TRUE, abbr = F)
ped_data$Time <- as.factor(ped_data$Time)
ped_data$WkYr <- paste(week(ped_data$Date_Time), year(ped_data$Date_Time), sep = "-")

@@

dfa$DayType <-       ifelse(dfa$Day == "Sunday", "Sunday",
                     ifelse(dfa$Day == "Saturday", "Saturday",
                    ifelse(dfa$Day == "Monday", "Monday",
                     ifelse(dfa$Day == "Friday", "Friday",
                                  "Midweek"))))
dfa$DayType[dfa$IsHDay == 1] <- "Holiday"
dfa$HDay <- ifelse(dfa$IsHDay == 1, "Holiday", dfa$Day)
dfa <- dfa %>% mutate(WkYr = paste(week(Date_Time), year(Date_Time)), sep = "-")
original_df <- dfa
dfa <- filter(dfa, Date_Time < "2017-01-01")
ori_2 <- filter(original_df, Date_Time < "2017-01-01")
```
# Why do we want to model pedestrian traffic?

@@

- Government example uses:  
  * Infrastructure planning
  * Security planning
- Private example uses:
  * Marketing campaign planning
  * Resource management
  * Investment planning
---
class: center
## The Problem: Prediction of pedestrian traffic
```{r, echo=FALSE, fig.retina = 3, fig.height=6, fig.width= 10}
p1 <- ggplot(dfa %>% filter(Date_Time > "2016-07-12 00:00:00",
                            Date_Time < "2016-07-25 00:00:00")) +
        geom_line(aes(x = Date_Time, y = `Flagstaff Station`)) +
        scale_y_continuous(limits = c(0, 7000))
p2 <- ggplot(dfa %>% filter(Date_Time > "2016-07-12 00:00:00",
                            Date_Time < "2016-07-25 00:00:00")) +
        geom_line(aes(x = Date_Time, y = `Southbank`)) +
        scale_y_continuous(limits = c(0, 7000))
p3 <- ggplot(dfa %>% filter(Date_Time > "2016-07-12 00:00:00",
                            Date_Time < "2016-07-25 00:00:00")) +
        geom_line(aes(x = Date_Time, y = `State Library`)) +
        scale_y_continuous(limits = c(0, 7000))
grid.arrange(p2, p1, p3, nrow = 3, ncol = 1)
```

--

#### Different locations have different counts and patterns

---
class: middle, center
# Missing Data
```{r, echo = FALSE, fig.retina=2, fig.width= 12, message=FALSE, warning=FALSE, cache = TRUE}
miss_plot_dat <- ped_data %>% filter(Date_Time > "2014-06-01",
                                     Date_Time < "2015-06-01") %>%
  group_by(WkYr, Sensor_ID) %>%
  summarise(Weekly_Count = sum(Hourly_Counts))

@@

random_sensors <- c("Collins Place (North)","Lygon St (West)","Spring St-Lonsdale St (South)","Flinders St-Spring St (West)","Convention/Exhibition Centre","Sandridge Bridge")

mdf <- miss_plot_dat2 %>% filter(Sensor_ID %in% random_sensors)

@@

---
class: middle
## Step 2c: Fit models at sensors with "large" proportion of missing values

@@

---
class: center, middle
## _Step 2c: Fit models at sensors with "large" proportion of missing values_

@@

p1 <- ggplot(tr_dat %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + theme_minimal() + ggtitle("Unscaled Neighbours") +
      theme(legend.position = "bottom")

p2 <- ggplot(tr_dat_sc %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + ggtitle("Scaled/Stadardised Neighbours") +
      ylab("Standard Deviation") + theme_minimal() +
      theme(legend.position = "none")
```

```{r, echo = FALSE, fig.retina = 6, message=FALSE, warning=F, fig.width=10, fig.height=6.5}
grid.arrange(p1, p2, ncol = 1)
```

>>>>

# Format of the data
#### Variables available:
- Time and date, sensor ID/location and hourly pedestrian count (volume)

```{r, echo=FALSE}
print(ped_data %>% as.tibble, n = 4)
```
```{r, include = FALSE, echo = FALSE, warning=FALSE, message=FALSE, cache = TRUE}
ped_data$Sensor_ID <- ped_data$Sensor
ped_data$Hourly_Counts <- ped_data$Count
ped_data$Date_Time <- ymd_h(paste(ped_data$Date, ped_data$Time, sep = " "))
ped_data$Day <- lubridate::wday(ped_data$Date_Time, label = TRUE, abbr = F)
ped_data$Month <- lubridate::month(ped_data$Date_Time, label = TRUE, abbr = F)
ped_data$Time <- as.factor(ped_data$Time)
ped_data$WkYr <- paste(week(ped_data$Date_Time), year(ped_data$Date_Time), sep = "-")

@@

dfa$DayType <-       ifelse(dfa$Day == "Sunday", "Sunday",
                     ifelse(dfa$Day == "Saturday", "Saturday",
                    ifelse(dfa$Day == "Monday", "Monday",
                     ifelse(dfa$Day == "Friday", "Friday",
                                  "Midweek"))))
dfa$DayType[dfa$IsHDay == 1] <- "Holiday"
dfa$HDay <- ifelse(dfa$IsHDay == 1, "Holiday", dfa$Day)
dfa <- dfa %>% mutate(WkYr = paste(week(Date_Time), year(Date_Time)), sep = "-")
original_df <- dfa
dfa <- filter(dfa, Date_Time < "2017-01-01")
ori_2 <- filter(original_df, Date_Time < "2017-01-01")
```

```{r, echo=FALSE, fig.retina = 3, fig.height=3, fig.width= 10, warning=F, message = F}
pd2 <- ggplot(dfa %>% 
              filter(Date_Time > "2016-07-12 00:00:00",
                     Date_Time < "2016-07-25 00:00:00")) +
        scale_y_continuous(limits = c(0, 7000)) +
        xlab("Time and Date") +
        theme_minimal()

p1 <-  pd2 + geom_line(aes(x = Date_Time, y = `Flagstaff Station`))
p2 <-  pd2 + geom_line(aes(x = Date_Time, y = `Southbank`))
p3 <-  pd2 + geom_line(aes(x = Date_Time, y = `State Library`))

grid.arrange(p1, p2, p3, ncol = 1)
```

---
class: center
# The Problem: Prediction of pedestrian traffic
```{r, echo=FALSE, fig.retina = 3, fig.height=5, fig.width= 10, warning=F, message = F}
pd <- ggplot(dfa %>% filter(Date_Time < "2017-01-01 00:00:00",
                            Date_Time > "2013-12-31 00:00:00")) +
        scale_y_continuous(limits = c(0, 7000)) +
        xlab(NULL) +
        theme_minimal()

p1 <-  pd + geom_line(aes(x = Date_Time, y = `Flagstaff Station`))
p2 <-  pd + geom_line(aes(x = Date_Time, y = `Southbank`))
p3 <-  pd + geom_line(aes(x = Date_Time, y = `State Library`)) +
            xlab("Time and Date")

grid.arrange(p2, p1, p3, nrow = 3, ncol = 1)
```

- Different locations have different counts and patterns
- We want to predict using 2014-2016 data to train our model

---
## Why do we want to model pedestrian traffic?

@@

- Government example uses:  
  * Infrastructure planning
  * Security planning
- Private example uses:
  * Marketing campaign planning
  * Resource management
  * Investment planning
---
class: middle, center
# Missing Data
```{r, echo = FALSE, fig.retina=2, fig.width= 12, message=FALSE, warning=FALSE, cache = TRUE}
miss_plot_dat <- ped_data %>% filter(Date_Time > "2014-06-01",
                                     Date_Time < "2015-06-01") %>%
  group_by(WkYr, Sensor_ID) %>%
  summarise(Weekly_Count = sum(Hourly_Counts))

@@

random_sensors <- c("Collins Place (North)","Southbank","Spring St-Lonsdale St (South)",
                    "Flinders St-Spring St (West)","Convention/Exhibition Centre","Sandridge Bridge")

mdf <- miss_plot_dat2 %>% filter(Sensor_ID %in% random_sensors) %>%
        mutate(Sensor_ID = as.factor(Sensor_ID))

@@

---
class: middle
## Step 2c: Find neighbours for sensors with "large" proportion of missing values

@@

---
class: center, middle
## _Step 2c: Fit find neighbours sensors with "large" proportion of missing values_

@@

p1 <- ggplot(tr_dat %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + theme_minimal() + ggtitle("Unscaled Neighbours") +
      theme(legend.position = "none")

p2 <- ggplot(tr_dat_sc %>% filter(Date_Time < "2017-03-02",
                         Date_Time > "2017-02-26")) +
      geom_line(aes(x = Date_Time, y = Count,
                    colour = Sensor), size = 2) +
      scale_colour_viridis(discrete = TRUE,
                           labels = c("Neighbour 1 (Flinders St Station Underpass)",
                                      "Neighbour 2 (Sandridge Bridge)",
                                      "Target (Southbank)")) +
      xlab("Time") + ggtitle("Scaled/Stadardised Neighbours") +
      ylab("Standard Deviation") + theme_minimal() +
      theme(legend.position = "bottom")
```

```{r, echo = FALSE, fig.retina = 6, message=FALSE, warning=F, fig.width=10, fig.height=6}
grid.arrange(p1, p2, ncol = 1)
```

- Need to standardise counts at neighbours to match patterns

---
## _Step 2c: Fit models using neighbours_
At each sensor location with "large" proportion of missing values, we estimate a generalised linear model with:

$$\mu_{\text{Time}, \text{HourlyCounts}_{\text{neighbours}}^{SC}} \sim \text{Time} \times \text{HourlyCounts}_{\text{neighbour 1}}^{SC} + \text{Time} \times \text{HourlyCounts}_{\text{neighbour 2}}^{SC}$$

and a quaispoisson error distribution.

- use interaction with `Time`, the hour of the day, and the scaled counts at neighbours
  * Again, `Time` is treated as a factor variable
  * Counts scaled by standardising each sensor to $\mu = 0$ and $\sigma^2 = 1$
